The hedging hyperprior for the normal distribution: unknown mean, known precision

### 2023-03-28 text/23/exponential-hedging-hyperprior.Rmd
Cleaning up hedging hyperprior formulas

You have an informative prior on $\theta$ that has an inverse gamma distribution

$g(\theta; \alpha, \beta)=\frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}$

and you collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of

$L(\mathbf{X})=\prod_{i=1}^n f(X_i)=\theta^{-n} e^{-n \bar X / \theta}$

@@

$h(\theta | \mathbf{X}) \propto \frac{\beta^\alpha}{\Gamma(\alpha)}\theta^{-\alpha-1}e^{-\beta / \theta}\theta^{-n} e^{-n \bar X / \theta}$

@@

$h(\theta | \mathbf{X}) \propto \theta^{-\alpha-1-n}e^{-(\beta + n\bar X) / \theta}$

which, thanks to the miracle of conjugancy, is an inverse gamma distribution. The posterior mean

$\frac{\beta + n\bar X}{\alpha-1+n}$

@@

$\Big(\frac{\alpha-1}{\alpha-1+n}\Big)\frac{\beta}{\alpha-1}+\Big(\frac{n}{\alpha-1+n}\Big)\bar X$

>>>>

You have an informative prior on $\theta$ that has an inverse gamma distribution with prior parameters $\alpha_0$ and $\beta_0$.

$g(\theta; \alpha_0, \beta_0) = 
  \frac
    {\beta^\alpha_0}
    {\Gamma(\alpha_0)}
  \theta^{-\alpha_0-1}
  e^{-\beta_0 / \theta}$

The mean of this prior distribution is

$\frac{\beta}{\alpha_0-1}$

You collect a sample of n data points, $\mathbf{X} = X_1, X_2, ..., X_n$. This gives you a likelihood of

$L(\mathbf{X})
  = \prod_{i=1}^n f(X_i)
  = \theta^{-n} e^{-n \bar X / \theta}$

@@

$h(\theta|\mathbf{X}) \propto 
  \frac
    {\beta_0^{\alpha_0-1}}
    {\Gamma(\alpha_0)}
  \theta^{-\alpha_0-1}
  e^{-\beta_0 / \theta}
  \theta^{-n}
  e^{-n \bar X / \theta}$

@@

$h(\theta | \mathbf{X}) \propto 
  \theta^{-\alpha_0-1-n}
  e^{-(\beta_0 + n\bar X) / \theta}$

which, thanks to the miracle of conjugancy, is an inverse gamma distribution. With posterior paramaters

$\alpha_1 = \alpha_0 + n$

$\beta_1 = \beta_0 + n\bar X$

The posterior mean

$\frac{\beta_0 + n\bar X}{\alpha_0-1+n}$

@@

$\Big(
  \frac
    {\alpha_0-1}
    {\alpha_0-1+n}\Big)
  \frac
    {\beta_0}
    {\alpha-1} + 
  \Big(
    \frac
      {n}
      {\alpha-1+n}\Big)
  \bar X$

### A simple numerical example

### 2023-03-28 text/23/normal-hedging-hyperprior-part-1.Rmd
Cleaning up hedging hyperprior formulas

@@

$\tau=\frac{1}{\sigma*2}$

The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance,

Assume that $\tau$ is known and equal to $\tau_F$. A more general case, with $\tau$ unknown is described on a separate web page. For simplicity, we will drop any multiplicative constants not involving X or $\mu$

@@

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0})^2}$

@@

$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau0}$

@@

This corresponds to a fairly informative prior distribution. This distribution places $\theta$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.

@@

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\theta$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.

@@

gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
density <- outer(mu_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  theta = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

>>>>

$f(X; \mu, \sigma) = \frac{1}{\sigma \sqrt{2 \pi}}e^{-(X-\mu)^2 / 2\sigma^2}$

@@

$\tau=\frac{1}{\sigma^2}$

The parameter $\tau$ is called the precision. Note that some define the precision differently as the inverse of the standard deviation rather than the inverse of the variance.

Assume that $\tau$ is known and equal to $\tau_F$. Other cases with [mean known and precision unknown][sim2] and with [both mean and precision unknown][sim3] appear on different web pages.

[sim2]: http://new.pmean.com/normal-hedging-hyperprior-part2
[sim3]: http://new.pmean.com/normal-hedging-hyperprior-part3

For simplicity, we will drop any multiplicative constants not involving X or $\mu$

@@

$h(\mu | \mathbf{X}) \propto e^{-(n\tau_F+\tau_0)(\mu-\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau_0})^2}$

@@

$\frac{n \tau_F \bar{X} + \tau_0 \mu_0}{n\tau_F+\tau_0}$

@@

This corresponds to a fairly informative prior distribution. This distribution places $\mu$ within `r sprintf("%.2f", prior05)` and `r sprintf("%.2f", prior95)` with 95% probability.

@@

The posterior distribution is not close to either the prior distribution or the likelihood. It places $\mu$ within `r sprintf("%.2f", post05)` and `r sprintf("%.2f", post95)` with 95% probability. It excludes the prior mean of `r mu0` and it excludes `r xbar`, the mean of the data.

@@

gstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF)+tauF)}
density <- outer(mu_grid, lambda_grid, gstar)
op <- par(bg = "white") 
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  mu = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

```{r}
density <- outer(mu_grid, lambda_grid, gstar)
density[16:44, ] <- NA
op <- par(bg = "white")
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  mu = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

fstar <- function(mu, lambda) {f(xbar, mu, tau0)^n}
likelihood <- outer(mu_grid, lambda_grid, fstar)
z_max <- max(likelihood, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  likelihood, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  mu = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```

@@

hstar <- function(mu, lambda) {g(mu, mu0, lambda*(tau0-tauF) + tauF)*f(xbar, mu, tauF)^n}
density <- outer(mu_grid, lambda_grid, hstar)
z_max <- max(density, na.rm=TRUE)
persp(
  mu_grid, 
  lambda_grid, 
  density, 
  zlim=c(0, 1.1*z_max),
  xlab="mu",
  ylab="lambda",
  mu = 135, 
  phi = 30, 
  expand=0.5,
  ticktype="detailed")
```
